Indoor Dining Shutdowns Negatively Affect Toronto Restaurants’ Performance and Business Practices

### 2021-02-26 outputs/paper/paper.Rmd
Edit for typos and grammar

---
title: "Indoor Dining Shutdowns Negatively Affect Toronto Restaurants’ Performance and Business Practices"
author: "Najma Osman, Rachit Srivastava, Morgaine Westin, and Bongju Yoo"
thanks: "Code and data are available at: https://github.com/bonjwow/covid-shut-down"
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "In light of recent COVID-19 restrictions and its impact on businesses, a randomized control trial was conducted by Petit Poll on behalf of the Ontario Government to test the effects of shutting down indoor dining services on restaurants in Toronto. Restaurants who received the intervention and only offered take-out or delivery services experienced a decline in their quarterly revenues compared to the previous quarter and had to make adjustments to the number of employees, menu prices, and operational schedules of their restaurants. Conversely, restaurants in the control group whose indoor dining services remained open during the study period experienced no significant changes in their revenue, employee count, menu prices, or operational schedule compared to the previous corner. While food delivery and take-out services are becoming more popular, these results indicate the value of indoor dining experiences for both restaurants and customers, which the Ontario Government can take into consideration when determining how they can best support Toronto’s growing restaurant industry. "
output:
  bookdown::pdf_document2:
    fig_caption: yes
geometry: margin=0.75in
header-includes:
- \usepackage{float} #use the 'float' package
- \floatplacement{figure}{H} #make every figure with caption = h
- \usepackage{longtable}
toc: FALSE
bibliography: references.bib
---

@@

Toronto is a cluster of diverse communities with many different aspects that make it a unique place to live. One of these aspects include restaurant businesses that consist of a wide variety of cuisines all across the city. The Ontario government department has asked us, Petit Poll, to design an experiment to test the implications of COVID-19 shutdowns and its effect on restaurant businesses. During Toronto’s tightest lockdown measures, many restaurants were forced to establish ghost kitchens in which their facilities were designed to focus exclusively on take-out and delivery orders [@kelly2019]. While some restaurants had already been using the ghost kitchen model before the pandemic, the circumstances of the pandemic undoubtedly lead to a rise in this practice [@Doradea2020]. Focusing on take-out and delivery services reduces the costs necessary to run a fully operational brick and mortar restaurant and removes the need for many customer-facing staff members [@kelly2019]. However, there are potential drawbacks that come with this shift as customers are losing the benefits and experiences of indoor dining  while restaurants are losing an important source of revenue. As pandemic restrictions loosen and restaurants are able to open up their indoor dining services again, it is possible that many restaurants may make the decision to continue to focus exclusively on take-out and delivery. Given the unprecedented nature of this situation, we do not yet fully understand the extent to which this shift towards take-out and delivery only services will impact restaurants and customer behaviours in the long term. This study aims to measure the effects on restaurant performance which may be associated with shutting down indoor dining, and the changes that restaurants make to deal with this shift.

As data scientists for Petit Poll, we narrowed down our test to restaurants in the city of Toronto as it condenses our test results into one precise area within Ontario. By using a business licensing dataset obtained from the Toronto Open Data Portal [@todp2021], we were able to obtain information on all restaurants in the city. Since eating establishments have been impacted by COVID-19 in the past year, to which there have been irregular restrictions on the businesses, we believed this characteristic was best suited for this study. In Toronto, pandemic restrictions likely had an impact on customers’ dining behaviours, as they were forced to order restaurant food for take-out and delivery only. Now that restaurants are able to open up their full services again, will their dining habits shift as they have become accustomed to the convenience of food delivery? Exploring how shutting down indoor dining services again after they have just reopened may affect performance such as revenue can also provide insights on Torontonians’ dining habits and preferences in these regards.

This study was conducted for three months, from October 1st to December 31st, 2020 where we studied the comparison between a control group and an intervention group. The intervention group had to partake in the study by resuming their business as a take-out only business and a control group had to remain business as usual. This was done intentionally to maintain the integrity of the study by studying the differences between restaurants that closed their indoor dining services and the ones that did not. Survey data was collected via Google Forms to measure the effects of the intervention on restaurant’s employee numbers and revenue, and if changes were made in menu prices and operation days and hours during the intervention period. Survey data were processed and analyzed in R [@citeR], with a complete list of packages and their usages listed in the Appendix. Based on survey responses, it was revealed that most restaurants in the intervention group who shut down their indoor dining services experienced a decrease in their total sales revenue compared to the previous quarter, with many reporting an overall decline of 50% or more. During the study period, restaurants in the intervention group made several changes to their restaurant practices to account for these changes such as reducing the number of employees, adjusting their menu prices, and reducing their hours and days of operation. Meanwhile, the restaurants in the control group who continued operating at full capacity remained relatively stable in terms of their revenue, and did not make any significant changes to their employee counts, menu prices, or operational periods. This suggests that shutdowns of indoor dining practices due to COVID-19 have large implications on restaurant performance, where the revenue, hours and days of operation, and menu prices all get impacted to adjust to deviations from the restaurant’s original business model.

@@

We implemented a randomized control trial to examine the effects of shutting down indoor dining services on restaurant performance. Survey methodology was used to measure these effects against a valid comparison group consisting of restaurants operating at regular capacity. In this report, we will refer to the systematic closure of indoor dining services and simultaneously switch to take-out or delivery only services as the intervention. The control group will refer to restaurants whose services remained fully operational for the duration of the study and who were invited to complete a survey regarding their restaurant’s performance after the intervention period ended.

A sample size of 400 was selected for the experiment based on funding availability (see Appendix for funding details), with 200 restaurants randomly assigned to receive the intervention and 200 restaurants randomly assigned to be in the control group. While these sample sizes indicate the number of restaurants we reached out, the number of respondents who consented to participate in the study differed slightly. A total of 150 restaurants received the intervention and responded to the survey, while 180 restaurants assigned to the control condition completed the survey.

@@

Restaurants in the intervention condition received $1200.00 CAD for participating. This amount was determined based on funding allocations for this study. We acknowledged that participating in the intervention may result in some financial losses and put some businesses at risk and wanted to provide some form of compensation to help alleviate those concerns as well as incentivize restaurants to participate. However, we also understand that this amount of money carries different weight for different types of restaurants. We provided resources and encouraged restaurants to look into external relief funding to support their businesses throughout and after the intervention period.

@@

The first three characters of the postal codes provided for each restaurant in the dataset were used to stratify each restaurant into boroughs assigned by Canada Post [@canpost]. Figure \@ref(fig:regions) displays the percentage distribution of restaurants located in each borough. As expected, a large proportion of restaurants were located in Downtown Toronto, followed by North York and Scarborough. After assigning each restaurant to an individual strata, a stratified random sampling method was implemented in R [@citeR] to randomly select a number of restaurants from each borough in proportion to the number of restaurants clustered in each area. For example, since our total sample size is 400 and approximately 28% of restaurants in the total list of restaurants we are sampling from are located in Downtown Toronto, we randomly selected 28% of 400 (112 restaurants) from the Downtown Toronto area to be in our sample. After restaurants were selected from each borough, R was used to randomly assign each restaurant to either the intervention group or the control group. The random assignment was done by borough, so half of the restaurants selected from each borough were assigned to the intervention group and the other half were assigned to the control. This sampling process ensured the restaurants we randomly selected were representative of all areas in Toronto, and that there were no major differences between restaurants assigned to the intervention group and those assigned to the control group in terms of their locations.

@@

* The largest concentration (nearly 60%) of restaurants in the Greater Toronto Area are centralized in downtown Toronto (28.26%), North York (15.78%), and Scarborough (13.74%); 
* Most restaurants are of medium or large scale^[The survey asked respondents to consider their restaurant scale in terms of space and seating capacity, however the perceptions of scale may be subjective]; 
* The majority of restaurants who responded are relatively new (Have been operating for 3 years or less); 
* Fast-food/quick service restaurants were the most common types, followed by casual dining, and then fine dining;
* During the intervention period, restaurants received the most orders via UberEats, followed by DoorDash, and then other food courier services;
* A small portion of restaurants utilized their own in-house delivery services, most of which were in the control group
* The distribution of these characteristics were roughly equal among restauraunts in both groups, meaning their sample characteristics are similar

@@

The subsection lists the R packages used in the making of this report.

>>>>

---
title: "Indoor Dining Shutdowns Negatively Affect Toronto Restaurants’ Performance and Business Practices"
author: "Najma Osman, Rachit Srivastava, Morgaine Westin, and Bongju Yoo"
thanks: "Code and data are available at: https://github.com/bonjwow/covid-shut-down"
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "In light of recent COVID-19 restrictions and its impact on businesses, a randomized control trial was conducted by Petit Poll on behalf of the Ontario Government to test the effects of shutting down indoor dining services on restaurants in Toronto. Restaurants who received the intervention and only offered take-out or delivery services experienced a decline in their revenue compared to the previous quarterly earnings, and had to make adjustments to the number of employees, menu prices, and operational schedules of their restaurants. Conversely, restaurants in the control group whose indoor dining services remained open during the study period experienced no significant changes in their revenue, employee count, menu prices, or operational schedule compared to the previous quarter. While food delivery and take-out services are becoming more popular, these results indicate the value of indoor dining experiences for both restaurants and customers, which the Ontario Government can take into consideration when determining how they can best support Toronto’s growing restaurant industry. "
output:
  bookdown::pdf_document2:
    fig_caption: yes
geometry: margin=0.75in
header-includes:
- \usepackage{float} #use the 'float' package
- \floatplacement{figure}{H} #make every figure with caption = h
- \usepackage{longtable}
toc: FALSE
bibliography: references.bib
---

@@

Toronto is a cluster of diverse communities with many different aspects that make it a unique place to live. One of these aspects include restaurant businesses that consist of a wide variety of cuisines all across the city. The Ontario government department has asked us, Petit Poll, to design an experiment to test the implications of COVID-19 shutdowns and its effect on restaurant businesses. During Toronto’s tightest lockdown measures, many restaurants were forced to establish ghost kitchens in which their facilities were designed to focus exclusively on take-out and delivery orders [@kelly2019]. While some restaurants had already been using the ghost kitchen model before the pandemic, the circumstances of the pandemic undoubtedly lead to a rise in this practice [@Doradea2020]. Focusing on take-out and delivery services reduces the costs necessary to run a fully operational brick and mortar restaurant and removes the need for many customer-facing staff members [@kelly2019]. However, there are potential drawbacks that come with this shift as customers are losing the benefits and experiences of indoor dining,  while restaurants are losing an important source of revenue. As pandemic restrictions loosen and restaurants are able to open up their indoor dining services again, it is possible that many restaurants may decide to continue to focus exclusively on take-out and delivery. Given the unprecedented nature of this situation, we do not yet fully understand the extent to which this shift towards take-out and delivery only services will impact restaurants and customer behaviours in the long term. This study aims to measure the effects on restaurant performance which may be associated with shutting down indoor dining, and the changes that restaurants make to deal with this shift.

As data scientists for Petit Poll, we narrowed down our test to restaurants in the city of Toronto as it condenses our test results into one precise area within Ontario. By using a business licensing dataset obtained from the Toronto Open Data Portal [@todp2021], we were able to obtain information on all restaurants in the city. Since eating establishments have been impacted by COVID-19 in the past year, to which there have been irregular restrictions on the businesses, we believed this characteristic was best suited for this study. In Toronto, pandemic restrictions likely had an impact on customers’ dining behaviours, as they were forced to order restaurant food for take-out and delivery only. Now that restaurants are able to open up their full services again, will customers' dining habits shift as they have become accustomed to the convenience of food delivery? Exploring how shutting down indoor dining services again after they have just reopened may affect performance such as revenue can also provide insights on Torontonians’ dining habits and preferences in these regards.

This study was conducted for three months, from October 1st to December 31st, 2020 where we studied the comparison between a control group and an intervention group. The intervention group had to partake in the study by their business as a take-out and delivery only restaurant, while the control group remained fully operational with indoor dining services available. This was done intentionally to maintain the integrity of the study by studying the differences between restaurants that closed their indoor dining services and the ones that did not. Survey data was collected via Google Forms to measure the effects of the intervention on restaurant’s employee numbers and revenue, and if changes were made in menu prices and operation days and hours during the intervention period. Survey data were processed and analyzed in R [@citeR], and a complete list of packages and their usages can be found in the Appendix. Based on survey responses, it was revealed that most restaurants in the intervention group who shut down their indoor dining services experienced a decrease in their total sales revenue compared to the previous quarter, with many reporting an overall decline of 50% or more. During the study period, restaurants in the intervention group made several changes to their restaurant practices to account for these changes such as reducing the number of employees, adjusting their menu prices, and reducing their hours and days of operation. Meanwhile, the restaurants in the control group who continued operating at full capacity remained relatively stable in terms of their revenue, and did not make any significant changes to their employee counts, menu prices, or operational periods. This suggests that shutdowns of indoor dining practices due to COVID-19 have large implications on restaurant performance, where the revenue, hours and days of operation, and menu prices all get impacted to adjust to deviations from the restaurant’s original business model.
```{r}

@@

We implemented a randomized control trial to examine the effects of shutting down indoor dining services on restaurant performance. Survey methodology was used to measure these effects against a valid comparison group consisting of restaurants operating at regular capacity. In this report, we will refer to the systematic closure of indoor dining services and subsequent switch to take-out or delivery only services as the intervention. The control group will refer to restaurants whose services remained fully operational for the duration of the study and who were invited to complete a survey regarding their restaurant’s performance after the intervention period ended.

A sample size of 400 was selected for the experiment based on funding availability (see Appendix for funding details), with 200 restaurants randomly assigned to receive the intervention and 200 restaurants randomly assigned to be in the control group. While these sample sizes indicate the number of restaurants we reached out to, the number of respondents who consented to participate in the study differed slightly. A total of 150 restaurants received the intervention and responded to the survey, while 180 restaurants assigned to the control condition completed the survey.

@@

Restaurants in the intervention condition received $1200.00 CAD for participating. This amount was determined based on funding allocations for this study. We acknowledged that participating in the intervention may result in some financial losses and put some businesses at risk, and wanted to provide some form of compensation to help alleviate those concerns as well as incentivize restaurants to participate. However, we also understand that this amount of money carries different weight for different types of restaurants. We provided resources and encouraged restaurants to look into external relief funding to support their businesses throughout and after the intervention period.

@@

The first three characters of the postal codes provided for each restaurant in the dataset were used to stratify each restaurant into boroughs assigned by Canada Post [@canpost]. Figure \@ref(fig:regions) displays the percentage distribution of restaurants located in each borough. As expected, a large proportion of restaurants were located in Downtown Toronto, followed by North York and Scarborough. After assigning each restaurant to an individual strata, a stratified random sampling method was implemented in R [@citeR] to randomly select a number of restaurants from each borough in proportion to the number of restaurants clustered in each area [@wu2020sampling]. For example, since our total sample size is 400 and approximately 28% of restaurants in the total list of restaurants we are sampling from are located in Downtown Toronto, we randomly selected 28% of 400 (112 restaurants) from the Downtown Toronto area to be in our sample. After restaurants were selected from each borough, R was used to randomly assign each restaurant to either the intervention group or the control group. The random assignment was done by borough, so half of the restaurants selected from each borough were assigned to the intervention group and the other half were assigned to the control. This sampling process ensured the restaurants we randomly selected were representative of all areas in Toronto, and that there were no major differences between restaurants assigned to the intervention group and those assigned to the control group in terms of their locations.

@@

* The largest concentration (nearly 60%) of restaurants in the Greater Toronto Area are centralized in downtown Toronto (28.26%), North York (15.78%), and Scarborough (13.74%)
* Most restaurants are of medium or large scale^[The survey asked respondents to consider their restaurant scale in terms of space and seating capacity, however the perceptions of scale may be subjective]
* The majority of restaurants who responded are relatively new (Have been operating for 3 years or less)
* Fast-food/quick service restaurants were the most common types, followed by casual dining, and then fine dining
* During the intervention period, restaurants received the most orders via UberEats, followed by DoorDash, and then other food courier services
* A small portion of restaurants utilized their own in-house delivery services, most of which were in the control group
* The distribution of these characteristics were roughly equal among restauraunts in both groups, meaning their sample characteristics are similar

@@

This subsection lists the R packages used in the making of this report.
